---
title: "NMDS Analysis Tutorial"
output:
  pdf_document: default
  html_document: default
---

## Introduction

This tutorial demonstrates how to perform a Non-Metric Dimensional Scaling (NMDS) analysis using R. We will use a sample dataset for better understanding.

## Setup

First, let's load the necessary packages.

```{r setup}
library(vegan)
library(tidyverse)
library(RColorBrewer)
```

## Sample Data Creation

To keep things simple, we are simulating a sample dataset for the tutorial.

```{r data-creation}
set.seed(123)  # Setting a seed for reproducibility

bsll <- data.frame(
  SampleNo = 1:20,
  Animal = sample(c("A", "B", "C", "D", "E"), 20, replace = TRUE),
  Phase = sample(c("PhaseA", "PhaseB"), 20, replace = TRUE),  # Renamed column & values
  Sp1 = rnorm(20, 10, 5),
  Sp2 = rnorm(20, 15, 4),
  Sp3 = rnorm(20, 12, 3)
)

bsll_env <- bsll %>% select(SampleNo, Animal, Phase)
bsll_sp <- bsll %>% select(-SampleNo, -Animal, -Phase)

```

## Data Preparation

Next, we prepare the data by converting necessary columns to factors and generating a colour palette.

```{r data-preparation}
bsll_env$Animal_factor <- as.factor(bsll_env$Animal)
num_sites <- length(unique(bsll_env$Animal_factor))
color_palette <- colorRampPalette(brewer.pal(9, "Set1"))(num_sites)
```

## NMDS Analysis

With the data ready, we can now compute the NMDS analysis.

```{r nmds-analysis}
NMDS_result <- metaMDS(bsll_sp, distance = "bray", k = 2)
```

## Plotting

Finally, let's visualize the NMDS results.

```{r plotting, fig.width=8, fig.height=6}
par(mar = c(5.1, 4.1, 4.1, 6))
plot(NMDS_result$points, 
     col=color_palette[as.integer(bsll_env$Animal_factor)], 
     pch=shape_vec, 
     cex=1.2, 
     main=paste("NMDS Bacterial Species Analysis\nStress:", round(NMDS_result$stress, 3)), 
     xlab= "NMDS Axis 1",
     ylab= "NMDS Axis 2")
ordispider(NMDS_result, groups=bsll_env$Phase, label = TRUE)
legend('topright',
       legend=levels(bsll_env$Animal_factor),
       fill=color_palette,
       title="Animal",
       cex=1,
       xpd = TRUE,
       inset = c(-0.10, 0))

```
